---
title: "AIC/BIC Logistic Regression Health"
author: "Michelle Le 30145965"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE)
library(dplyr) # data manipulation
library(naniar) # visualizing non-response
library(VIM) # visualizing non-response
library(mice) # multiple imputation
library(naniar) # summarize missingness patterns
library(ggplot2) # general data visualization
library(patchwork) # formatting data visualization
library(poLCA) # lca modelling
library(survey)
library(car)
```

# Load Data
```{r load data}
data <- read.csv("/Users/michellele/Downloads/data 501/5012-NGS-END_2020-CSV/ngs2020_pumf.csv", colClasses = "character")

# filter the data to bachelors only
data <- data %>%
  filter(CERTLEVP == 2) # CERTLEVP refers to university certificate or diploma below a bachelor's degree AND bachelor's degree
```

# Load the Imputed Datasets

```{r load the imputed datasets}
# load the imputed datasets
data_completed_1 <- read.csv("/Users/michellele/Downloads/data 501/data_completed_imp1.csv", colClasses = "character")
data_completed_1$PUMFID <- data$PUMFID
data_completed_1$LFSTATP <- data$LFSTATP

data_completed_2 <- read.csv("/Users/michellele/Downloads/data 501/data_completed_imp2.csv", colClasses = "character")
data_completed_2$PUMFID <- data$PUMFID
data_completed_2$LFSTATP <- data$LFSTATP

data_completed_3 <- read.csv("/Users/michellele/Downloads/data 501/data_completed_imp3.csv", colClasses = "character")
data_completed_3$PUMFID <- data$PUMFID
data_completed_3$LFSTATP <- data$LFSTATP

data_completed_4 <- read.csv("/Users/michellele/Downloads/data 501/data_completed_imp4.csv", colClasses = "character")
data_completed_4$PUMFID <- data$PUMFID
data_completed_4$LFSTATP <- data$LFSTATP

data_completed_5 <- read.csv("/Users/michellele/Downloads/data 501/data_completed_imp5.csv", colClasses = "character")
data_completed_5$PUMFID <- data$PUMFID
data_completed_5$LFSTATP <- data$LFSTATP

# convert to factor
data_completed_1[] <- lapply(data_completed_1, function(x) {
  if (is.character(x)) as.factor(x) else x
})
# str(data_completed_1)

data_completed_2[] <- lapply(data_completed_2, function(x) {
  if (is.character(x)) as.factor(x) else x
})
# str(data_completed_2)

data_completed_3[] <- lapply(data_completed_3, function(x) {
  if (is.character(x)) as.factor(x) else x
})
# str(data_completed_3)

data_completed_4[] <- lapply(data_completed_4, function(x) {
  if (is.character(x)) as.factor(x) else x
})
# str(data_completed_4)

data_completed_5[] <- lapply(data_completed_5, function(x) {
  if (is.character(x)) as.factor(x) else x
})
# str(data_completed_5)

# filter out rows where LFSTATP != 1, 2 (only want employed and unemployed status)
data_filtered_1 <- data_completed_1[data_completed_1$LFSTATP %in% c(1, 2), ]
data_filtered_2 <- data_completed_2[data_completed_2$LFSTATP %in% c(1, 2), ]
data_filtered_3 <- data_completed_3[data_completed_3$LFSTATP %in% c(1, 2), ]
data_filtered_4 <- data_completed_4[data_completed_4$LFSTATP %in% c(1, 2), ]
data_filtered_5 <- data_completed_5[data_completed_5$LFSTATP %in% c(1, 2), ]

# Filtering for program 9: Health
data_filtered_1 = filter(data_filtered_1, PGMCIPAP == 9)
data_filtered_2 = filter(data_filtered_2, PGMCIPAP == 9)
data_filtered_3 = filter(data_filtered_3, PGMCIPAP == 9)
data_filtered_4 = filter(data_filtered_4, PGMCIPAP == 9)
data_filtered_5 = filter(data_filtered_5, PGMCIPAP == 9)

dim(data_filtered_1)
dim(data_filtered_2)
dim(data_filtered_3)
dim(data_filtered_4)
dim(data_filtered_5)

data_filtered_1$LFSTATP = ifelse(data_filtered_1$LFSTATP == 2,0,1) # unemployed = 0; employed = 1
data_filtered_2$LFSTATP = ifelse(data_filtered_2$LFSTATP == 2,0,1)
data_filtered_3$LFSTATP = ifelse(data_filtered_3$LFSTATP == 2,0,1)
data_filtered_4$LFSTATP = ifelse(data_filtered_4$LFSTATP == 2,0,1)
data_filtered_5$LFSTATP = ifelse(data_filtered_5$LFSTATP == 2,0,1)
```

# Weights set-up

This section adds back the weights to their corresponding imputed dataset based on PUMFID.

```{r add back weights}
data_filtered_1 <- data_filtered_1 %>%
  left_join(
    dplyr::select(data, PUMFID, WTPF),
    by = c("PUMFID" = "PUMFID")
  )

data_filtered_2 <- data_filtered_2 %>%
  left_join(
    dplyr::select(data, PUMFID, WTPF),
    by = c("PUMFID" = "PUMFID")
  )

data_filtered_3 <- data_filtered_3 %>%
  left_join(
    dplyr::select(data, PUMFID, WTPF),
    by = c("PUMFID" = "PUMFID")
  )

data_filtered_4 <- data_filtered_4 %>%
  left_join(
    dplyr::select(data, PUMFID, WTPF),
    by = c("PUMFID" = "PUMFID")
  )

data_filtered_5 <- data_filtered_5 %>%
  left_join(
    dplyr::select(data, PUMFID, WTPF),
    by = c("PUMFID" = "PUMFID")
  )

# Converting WTPF as numeric
data_filtered_1$WTPF = as.numeric(data_filtered_1$WTPF)
data_filtered_2$WTPF = as.numeric(data_filtered_2$WTPF) 
data_filtered_3$WTPF = as.numeric(data_filtered_3$WTPF)
data_filtered_4$WTPF = as.numeric(data_filtered_4$WTPF)
data_filtered_5$WTPF = as.numeric(data_filtered_5$WTPF)
```

# Training and Testing Split

Create training (70%) and test (30%) sets.

```{r training and test split}
# Creating training and test sets
set.seed(1)
train <- sample(nrow(data_filtered_1), 0.7*nrow(data_filtered_1), replace = FALSE)

train_1 <- data_filtered_1[train,]
train_2 <- data_filtered_2[train,]
train_3 <- data_filtered_3[train,]
train_4 <- data_filtered_4[train,]
train_5 <- data_filtered_5[train,]


test_1 <- data_filtered_1[-train,]
test_2 <- data_filtered_2[-train,]
test_3 <- data_filtered_3[-train,]
test_4 <- data_filtered_4[-train,]
test_5 <- data_filtered_5[-train,]
```

# Prepare Data for Stepwise AIC/BIC Logistic Regression

```{r prepare for stepwise AIC/BIC logistic regression modeling}
predictors_1 <- setdiff(names(data_filtered_1), c("LFCINDP", "PUMFID", "LFSTATP", "WTPF", "PGMCIPAP"))
predictors_2 <- setdiff(names(data_filtered_2), c("LFCINDP", "PUMFID", "LFSTATP", "WTPF", "PGMCIPAP"))
predictors_3 <- setdiff(names(data_filtered_3), c("LFCINDP", "PUMFID", "LFSTATP", "WTPF", "PGMCIPAP"))
predictors_4 <- setdiff(names(data_filtered_4), c("LFCINDP", "PUMFID", "LFSTATP", "WTPF", "PGMCIPAP"))
predictors_5 <- setdiff(names(data_filtered_5), c("LFCINDP", "PUMFID", "LFSTATP", "WTPF", "PGMCIPAP"))

formula_1 <- as.formula(
  paste("LFSTATP ~", paste(predictors_1, collapse = " + "))
)
formula_2 <- as.formula(
  paste("LFSTATP ~", paste(predictors_2, collapse = " + "))
)
formula_3 <- as.formula(
  paste("LFSTATP ~", paste(predictors_3, collapse = " + "))
)
formula_4 <- as.formula(
  paste("LFSTATP ~", paste(predictors_4, collapse = " + "))
)
formula_5 <- as.formula(
  paste("LFSTATP ~", paste(predictors_5, collapse = " + "))
)

full_model_1 <- glm(formula_1,
             data = data_filtered_1, subset = train,
             family = binomial)
full_model_2 <- glm(formula_2,
             data = data_filtered_2, subset = train,
             family = binomial)
full_model_3 <- glm(formula_3,
             data = data_filtered_3, subset = train,
             family = binomial)
full_model_4 <- glm(formula_4,
             data = data_filtered_4, subset = train,
             family = binomial)
full_model_5 <- glm(formula_5,
             data = data_filtered_5, subset = train,
             family = binomial)
```

```{r stepwise aic logistic regression}
aic_step_model_1 <- step(full_model_1, direction = "both", trace = 0)
# summary(aic_step_model_1)

aic_step_model_2 <- step(full_model_2, direction = "both", trace = 0)
# summary(aic_step_model_2)

aic_step_model_3 <- step(full_model_3, direction = "both", trace = 0)
# summary(aic_step_model_3)

aic_step_model_4 <- step(full_model_4, direction = "both", trace = 0)
# summary(aic_step_model_4)

aic_step_model_5 <- step(full_model_5, direction = "both", trace = 0)
# summary(aic_step_model_5)
```

# Check Assumptions for AIC Model

```{r check assumptions for stepwise aic models}
# Assumption 1: No severe multicollinearity
# vif(aic_step_model_1)
# vif(aic_step_model_2)
# vif(aic_step_model_3)
# vif(aic_step_model_4)
# vif(aic_step_model_5)

vif_results_aic_1 <- tryCatch({
  car::vif(aic_step_model_1)
}, error = function(e) {
  # This code runs if an error occurs
  return("VIF not calculated: Model contains fewer than 2 terms (or other error).")
})
print(vif_results_aic_1)

vif_results_aic_2 <- tryCatch({
  car::vif(aic_step_model_2)
}, error = function(e) {
  # This code runs if an error occurs
  return("VIF not calculated: Model contains fewer than 2 terms (or other error).")
})
print(vif_results_aic_2)

vif_results_aic_3 <- tryCatch({
  car::vif(aic_step_model_3)
}, error = function(e) {
  # This code runs if an error occurs
  return("VIF not calculated: Model contains fewer than 2 terms (or other error).")
})
print(vif_results_aic_3)

vif_results_aic_4 <- tryCatch({
  car::vif(aic_step_model_4)
}, error = function(e) {
  # This code runs if an error occurs
  return("VIF not calculated: Model contains fewer than 2 terms (or other error).")
})
print(vif_results_aic_4)

vif_results_aic_5 <- tryCatch({
  car::vif(aic_step_model_5)
}, error = function(e) {
  # This code runs if an error occurs
  return("VIF not calculated: Model contains fewer than 2 terms (or other error).")
})
print(vif_results_aic_5)

# Assumption 2: Checking for influential points using Cook's Distance
cooks_1 <- which(cooks.distance(aic_step_model_1) > 4/nrow(data_filtered_1[train,]))
# cooks_1
cooks_2 <- which(cooks.distance(aic_step_model_2) > 4/nrow(data_filtered_2[train,]))
# cooks_2
cooks_3 <- which(cooks.distance(aic_step_model_3) > 4/nrow(data_filtered_3[train,]))
# cooks_3
cooks_4 <- which(cooks.distance(aic_step_model_4) > 4/nrow(data_filtered_4[train,]))
# cooks_4
cooks_5 <- which(cooks.distance(aic_step_model_5) > 4/nrow(data_filtered_5[train,]))
# cooks_5
```

```{r stepwise bic logistic regression}
bic_step_model_1 <- step(full_model_1, k = log(nrow(data_filtered_1)), trace = 0)
# summary(bic_step_model_1)

bic_step_model_2 <- step(full_model_2, k = log(nrow(data_filtered_2)), trace = 0)
# summary(bic_step_model_2)

bic_step_model_3 <- step(full_model_3, k = log(nrow(data_filtered_3)), trace = 0)
# summary(bic_step_model_3)

bic_step_model_4 <- step(full_model_4, k = log(nrow(data_filtered_4)), trace = 0)
# summary(bic_step_model_4)

bic_step_model_5 <- step(full_model_5, k = log(nrow(data_filtered_5)), trace = 0)
# summary(bic_step_model_5)
```

# Check Assumptions for BIC Model

```{r check assumptions for stepwise bic models}
# Assumption 1: No severe multicollinearity
# vif(bic_step_model_1)
# vif(bic_step_model_2)
# vif(bic_step_model_3)
# vif(bic_step_model_4)
# vif(bic_step_model_5)

vif_results_bic_1 <- tryCatch({
  car::vif(bic_step_model_1)
}, error = function(e) {
  # This code runs if an error occurs
  return("VIF not calculated: Model contains fewer than 2 terms (or other error).")
})
print(vif_results_bic_1)

vif_results_bic_2 <- tryCatch({
  car::vif(bic_step_model_2)
}, error = function(e) {
  # This code runs if an error occurs
  return("VIF not calculated: Model contains fewer than 2 terms (or other error).")
})
print(vif_results_bic_2)

vif_results_bic_3 <- tryCatch({
  car::vif(bic_step_model_3)
}, error = function(e) {
  # This code runs if an error occurs
  return("VIF not calculated: Model contains fewer than 2 terms (or other error).")
})
print(vif_results_bic_3)

vif_results_bic_4 <- tryCatch({
  car::vif(bic_step_model_4)
}, error = function(e) {
  # This code runs if an error occurs
  return("VIF not calculated: Model contains fewer than 2 terms (or other error).")
})
print(vif_results_bic_4)

vif_results_bic_5 <- tryCatch({
  car::vif(bic_step_model_5)
}, error = function(e) {
  # This code runs if an error occurs
  return("VIF not calculated: Model contains fewer than 2 terms (or other error).")
})
print(vif_results_bic_5)

# Assumption 2: Checking for influential points using Cook's Distance
cooks_1 <- which(cooks.distance(bic_step_model_1) > 4/nrow(data_filtered_1[train,]))
# cooks_1
cooks_2 <- which(cooks.distance(bic_step_model_2) > 4/nrow(data_filtered_2[train,]))
# cooks_2
cooks_3 <- which(cooks.distance(bic_step_model_3) > 4/nrow(data_filtered_3[train,]))
# cooks_3
cooks_4 <- which(cooks.distance(bic_step_model_4) > 4/nrow(data_filtered_4[train,]))
# cooks_4
cooks_5 <- which(cooks.distance(bic_step_model_5) > 4/nrow(data_filtered_5[train,]))
# cooks_5
```

# Check for Common AIC Predictors

Predictors that are common amongst all 5 imputated datasets are meaningful.

```{r check common aic predictors}
aic_vars_1 <- setdiff(names(coef(aic_step_model_1)), "(Intercept)")
aic_vars_2 <- setdiff(names(coef(aic_step_model_2)), "(Intercept)")
aic_vars_3 <- setdiff(names(coef(aic_step_model_3)), "(Intercept)")
aic_vars_4 <- setdiff(names(coef(aic_step_model_4)), "(Intercept)")
aic_vars_5 <- setdiff(names(coef(aic_step_model_5)), "(Intercept)")

all_aic_vars <- c(aic_vars_1, aic_vars_2, aic_vars_3, aic_vars_4, aic_vars_5)
aic_tab <- table(all_aic_vars)
common_aic_vars <- names(aic_tab[aic_tab == 5])
common_aic_vars
```

# Check for Common BIC Predictors

Predictors that are common amongst all 5 imputated datasets are meaningful.

```{r check common bic predictors}
bic_vars_1 <- setdiff(names(coef(bic_step_model_1)), "(Intercept)")
bic_vars_2 <- setdiff(names(coef(bic_step_model_2)), "(Intercept)")
bic_vars_3 <- setdiff(names(coef(bic_step_model_3)), "(Intercept)")
bic_vars_4 <- setdiff(names(coef(bic_step_model_4)), "(Intercept)")
bic_vars_5 <- setdiff(names(coef(bic_step_model_5)), "(Intercept)")

all_bic_vars <- c(bic_vars_1, bic_vars_2, bic_vars_3, bic_vars_4, bic_vars_5)
bic_tab <- table(all_bic_vars)
common_bic_vars <- names(bic_tab[bic_tab == 5])
common_bic_vars
common_bic_vars <- names(bic_tab[bic_tab == 4])
common_bic_vars
common_bic_vars <- names(bic_tab[bic_tab == 3])
common_bic_vars
```
